---
title: "Quantifying and correcting slide-to-slide variation in multiplexed immunofluorescence images (analysis methods and code)"
author: "Coleman Harris, coleman.r.harris@vanderbilt.edu"
date: "2021-07-09"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
---

This notebook details the various analyses and code implementations used in the paper, "Quantifying and correcting slide-to-slide variation in multiplexed immunofluorescence images." Please see [here](https://colemanrharris.me) for more information on the paper and for complete descriptions of the analyses and concepts therein. The analyses here rely also on the following scripts, available in the Github repository, for further functionality:

- `all_combat_functions_210709.R`: code to run the ComBat algorithm
- `all_fda_functions_210709.R`: code to run the functional data registration algorithm
- `all_otsu_functions_210709.R`: code to run collect Otsu threshold and metrics

## Setup

```{r setup,results='hide',message=FALSE,warning=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE,fig.width = 10,message = FALSE)
## tidyverse packages
require(dplyr)
require(tidyr)

## visualization packages
require(ggplot2)
require(ggpubr)
require(ggforce)
require(viridis)

## methods packages
require(fda)
require(umap)
require(lme4)
require(emmeans)
require(uwot)
require(reticulate)

## set useful variables for later on
x_axis_text_size = 8
vals = rev(inferno(43))
```

## Normalization analyses
### Load data

```{r load data}
## load data from https://github.com/statimagcoll/atlasAnalysis
sc = atlasAnalysis::loadQuantifiedPCA()

## remove other cols that aren't median cell intensities
to_remove = grep("Median_Cyt|Median_Nuc|Median_Mem",colnames(sc))
sc = sc[,-to_remove]

## select markers for analysis
all_vars = c("Median_Cell_CD3D","Median_Cell_CD8","Median_Cell_COLLAGEN",
             "Median_Cell_VIMENTIN","Median_Cell_PANCK","Median_Cell_NAKATPASE",
             "Median_Cell_SOX9","Median_Cell_BCATENIN","Median_Cell_OLFM4")

## -- set all methods
methods = c("log10","mean_division","mean_division_log10","cube_root")
methods = c(methods,paste0(methods[1:4],"_fda_registered_x1"))
methods = c(methods,paste0(methods[1:4],"_ComBat_incl_zeroes"))
## -- !end set all methods
```

### Run transformations
Run the scale transformations on each variable as selected above. The transformations included are as follows:

- log10: $\log_{10}(y + 1)$
- cube root: $\sqrt[3]{y}$
- mean division: $\frac{y}{\mu_{ic}}$
- mean division log10: $\log_{10}(\frac{y}{\mu_{ic}} + \frac{1}{2})$

```{r transforms}
## --- --- ---
## --- log10 + 1
## --- --- ---
for(v in all_vars){
  sc[,paste0(v,"_log10")] = log10(sc[,v]+1)
}

## --- --- ---
## --- cube root
## --- --- ---
for(v in all_vars){
  sc[,paste0(v,"_cube_root")] = sc[,v]^(1/3)
}

## --- --- ---
## --- mean division, log mean division
## --- --- ---
for(v in all_vars){
	slides = split(sc,sc$SlideID)
	joe_l = list()
	for(i in 1:length(slides)){
	    s = data.frame(slides[[i]])
	    s[,paste0(v,"_mean_division")] = s[,v] / mean(s[,v],na.rm=TRUE)
	    joe_l[[i]] = s
	}
	sc = data.frame(data.table::rbindlist(joe_l))
	sc[,paste0(v,"_mean_division_log10")] = log10(sc[,paste0(v,"_mean_division")]+0.5)
	
	sc[,paste0(v,"_mean_division")] = sc[,paste0(v,"_mean_division")] + -min(sc[,paste0(v,"_mean_division")])
	sc[,paste0(v,"_mean_division_log10")] = sc[,paste0(v,"_mean_division_log10")] + -min(sc[,paste0(v,"_mean_division_log10")])
}
## --- --- ---
## --- rename raw values for later analysis
## --- --- ---
for(v in all_vars){
  sc[,paste0(v,"_raw")] = sc[,v]
}
```

### Run the functional data registration algorithm

```{r fda,results='hide'}
## --- --- ---
## --- fda
## --- --- ---
source("all_fda_functions_210709.R")

## loop over all variables and methods to run the registration
for(j in all_vars){
  for(i in methods[1:4]){
    ## update variable names
    normedVar = paste0(j,"_",i,"_fda_registered_x")
    var = paste0(j,"_",i)
    
    ## run 1 iteration of registration
    for(k in 1){
      sc = register_var(var = var,
                        normedVar = paste0(normedVar,k),
                        normedVariter = k,
                        cb_atl = sc)
      var = paste0(normedVar,k)
    }
  }
}
```

### Run the ComBat algorithm

```{r combat}
## --- --- ---
## --- ComBat
## --- --- ---
source("all_combat_functions_210709.R")

## loop over all variables and methods to run ComBat
for(j in all_vars){
  for(m in methods[1:4]){
    ## update variable name
    scale_channel = paste0(j,"_",m)
    
    ## run ComBat algorithm on given method and marker channel
    sc[,paste0(scale_channel,"_ComBat_incl_zeroes")] = adjust_vals(channel = scale_channel,slide_var="SlideID",chan=sc,remove_zeroes = FALSE)
  }
}
```

### Run evaluation metrics for downstream analysis
```{r otsus}
## --- --- ---
## --- Otsu thresholds
## --- --- ---
source("all_otsu_functions_210709.R")

## list for adjusting variable names
adjList = list("log10" ="Unnormalized: log10",
  "mean_division" ="Unnormalized: Mean division",
  "mean_division_log10" ="Unnormalized: Mean division log10",
  "cube_root" ="Unnormalized: Cube root",
  "log10_fda_registered_x1" ="fda: log10",
  "mean_division_fda_registered_x1" ="fda: Mean division",
  "mean_division_log10_fda_registered_x1" ="fda: Mean division log10",
  "cube_root_fda_registered_x1" ="fda: Cube root",
  "log10_ComBat_incl_zeroes" ="ComBat: log10",
  "mean_division_ComBat_incl_zeroes" ="ComBat: Mean division",
  "mean_division_log10_ComBat_incl_zeroes" ="ComBat: Mean division log10",
  "cube_root_ComBat_incl_zeroes" ="ComBat: Cube root")

## --- generate dataset of Otsu thresholds
otsus  = generate_otsu_data(sc,all_vars = all_vars,c(methods,"raw"))

## --- run misclassification metrics on Otsus
all_means = run_misclassification(sc,otsus,c(methods,"raw"))
```

## Paper figures
### Figure 1: Visual comparison of vimentin marker densities for each transformation method
```{r fig1,fig.height=7}
htan1 = ggplot(sc) +
  geom_density(aes(x=Median_Cell_VIMENTIN_log10,color=SlideID)) + 
  geom_rug(data=otsus[otsus$channel=="Median_Cell_VIMENTIN",],aes(x=log10_threshold,color=SlideID)) +
  theme(legend.position = "None") +
  scale_x_continuous(name=  paste0("log10"),limits = c(0,5)) +
  scale_y_continuous(name="Density",limits=c(0,2)) +
  scale_color_manual(values = vals) +
  theme(axis.title.x = element_text(size=x_axis_text_size))

htan2 = ggplot(sc) +
  geom_density(aes(x=Median_Cell_VIMENTIN_log10_ComBat_incl_zeroes,color=SlideID)) + 
  geom_rug(data=otsus[otsus$channel=="Median_Cell_VIMENTIN",],aes(x=log10_ComBat_incl_zeroes_threshold,color=SlideID)) +
  theme(legend.position = "None") +
  scale_x_continuous(name=  paste0("log10 (ComBat)")) +
  scale_y_continuous(name="Density") +
  scale_color_manual(values = vals) +
  theme(axis.title.x = element_text(size=x_axis_text_size))
  
htan3 = ggplot(sc) +
  geom_density(aes(x=Median_Cell_VIMENTIN_log10_fda_registered_x1,color=SlideID)) + 
  geom_rug(data=otsus[otsus$channel=="Median_Cell_VIMENTIN",],aes(x=log10_fda_registered_x1_threshold,color=SlideID)) +
  theme(legend.position = "None") +
  scale_x_continuous(name=  paste0("log10 (fda)"),limits = c(0,5)) +
  scale_y_continuous(name="Density",limits=c(0,2)) +
  scale_color_manual(values = vals) +
  theme(axis.title.x = element_text(size=x_axis_text_size))

htan4 = ggplot(sc) +
  geom_density(aes(x=Median_Cell_VIMENTIN_cube_root,color=SlideID)) + 
  geom_rug(data=otsus[otsus$channel=="Median_Cell_VIMENTIN",],aes(x=cube_root_threshold,color=SlideID)) +
  theme(legend.position = "None") +
  scale_x_continuous(name=  paste0("Cube root"),limits=c(0,30)) +
  scale_y_continuous(name="Density") +
  scale_color_manual(values = vals) +
  theme(axis.title.x = element_text(size=x_axis_text_size))

htan5 = ggplot(sc) +
  geom_density(aes(x=Median_Cell_VIMENTIN_cube_root_ComBat_incl_zeroes,color=SlideID)) + 
  geom_rug(data=otsus[otsus$channel=="Median_Cell_VIMENTIN",],aes(x=cube_root_ComBat_incl_zeroes_threshold,color=SlideID)) +
  theme(legend.position = "None") +
  scale_x_continuous(name=  paste0("Cube root (ComBat)"),limits=c(-25,25)) +
  scale_y_continuous(name="Density") +
  scale_color_manual(values = vals) +
  theme(axis.title.x = element_text(size=x_axis_text_size))

htan6 = ggplot(sc) +
  geom_density(aes(x=Median_Cell_VIMENTIN_cube_root_fda_registered_x1,color=SlideID)) + 
  geom_rug(data=otsus[otsus$channel=="Median_Cell_VIMENTIN",],aes(x=cube_root_fda_registered_x1_threshold,color=SlideID)) +
  theme(legend.position = "None") +
  scale_x_continuous(name=  paste0("Cube root (fda) "),limits=c(0,20)) +
  scale_y_continuous(name="Density") +
  scale_color_manual(values = vals) +
  theme(axis.title.x = element_text(size=x_axis_text_size))

htan7 = ggplot(sc) +
  geom_density(aes(x=Median_Cell_VIMENTIN_mean_division,color=SlideID)) + 
  geom_rug(data=otsus[otsus$channel=="Median_Cell_VIMENTIN",],aes(x=mean_division_threshold,color=SlideID)) +
  theme(legend.position = "None") +
  scale_x_continuous(name=  paste0("Mean division"),limits=c(0,5)) +
  scale_y_continuous(name="Density",limits=c(0,3)) +
  scale_color_manual(values = vals) +
  theme(axis.title.x = element_text(size=x_axis_text_size))

htan8 = ggplot(sc) +
  geom_density(aes(x=Median_Cell_VIMENTIN_mean_division_ComBat_incl_zeroes,color=SlideID)) + 
  geom_rug(data=otsus[otsus$channel=="Median_Cell_VIMENTIN",],aes(x=mean_division_ComBat_incl_zeroes_threshold,color=SlideID)) +
  theme(legend.position = "None") +
  scale_x_continuous(name=  paste0("Mean division (ComBat)"),limits=c(0,2)) +
  scale_y_continuous(name="Density") +
  scale_color_manual(values = vals) +
  theme(axis.title.x = element_text(size=x_axis_text_size))

htan9 = ggplot(sc) +
    geom_density(aes(x=Median_Cell_VIMENTIN_mean_division_fda_registered_x1,color=SlideID)) + 
  geom_rug(data=otsus[otsus$channel=="Median_Cell_VIMENTIN",],aes(x=mean_division_fda_registered_x1_threshold,color=SlideID)) +
    theme(legend.position = "None") +
    scale_x_continuous(name=  paste0("Mean division"),limits=c(0,5)) +
    scale_y_continuous(name="Density",limits=c(0,3)) +
    scale_color_manual(values = vals) +
    theme(axis.title.x = element_text(size=x_axis_text_size))

htan10 = ggplot(sc) +
  geom_density(aes(x=Median_Cell_VIMENTIN_mean_division_log10,color=SlideID)) + 
  geom_rug(data=otsus[otsus$channel=="Median_Cell_VIMENTIN",],aes(x=mean_division_log10_threshold,color=SlideID)) +
  theme(legend.position = "None") +
  scale_x_continuous(name=  paste0("Mean division log10"),limits=c(0,2)) +
  scale_y_continuous(name="Density",limits=c(0,5)) +
  scale_color_manual(values = vals) +
  theme(axis.title.x = element_text(size=x_axis_text_size))

htan11 = ggplot(sc) +
  geom_density(aes(x=Median_Cell_VIMENTIN_mean_division_log10_ComBat_incl_zeroes,color=SlideID)) + 
  geom_rug(data=otsus[otsus$channel=="Median_Cell_VIMENTIN",],aes(x=mean_division_log10_ComBat_incl_zeroes_threshold,color=SlideID)) +
  theme(legend.position = "None") +
  scale_x_continuous(name=  paste0("Mean division log10 (ComBat)")) +
  scale_y_continuous(name="Density") +
  scale_color_manual(values = vals) +
  theme(axis.title.x = element_text(size=x_axis_text_size))

htan12 = ggplot(sc) +
  geom_density(aes(x=Median_Cell_VIMENTIN_mean_division_log10_fda_registered_x1,color=SlideID)) + 
  geom_rug(data=otsus[otsus$channel=="Median_Cell_VIMENTIN",],aes(x=mean_division_log10_fda_registered_x1_threshold,color=SlideID)) +
  theme(legend.position = "None") +
  scale_x_continuous(name=  paste0("Mean division log10 (fda)")) +
  scale_y_continuous(name="Density") +
  scale_color_manual(values = vals) +
  theme(axis.title.x = element_text(size=x_axis_text_size))

## fig1 plot
ggpubr::ggarrange(htan1,htan2,htan3,
                  htan4,htan5,htan6,
                  htan7,htan8,htan9,
                  htan10,htan11,htan12,nrow=4,ncol =3)
```

### Figure 2A: Otsu misclassification
```{r fig2a,fig.height=6}
## --- calculate as ratios to raw data
all_ratios=all_means
all_ratios[,2:ncol(all_ratios)] = round(matrix(all_means[,2:ncol(all_ratios)]) - as.vector(data.frame(all_means[13,2:ncol(all_ratios)])),3)

not_rows = c("raw")
plot_values = all_ratios[!(all_ratios$method %in% not_rows),]

## change names for cleaner plotting
method_values = c("Unnormalized: Cube root",
                       "ComBat: Cube root", 
                       "fda: Cube root",
                       "Unnormalized: log10",
                       "ComBat: log10", 
                       "fda: log10",
                       "Unnormalized: Mean division",
                       "ComBat: Mean division", 
                       "fda: Mean division",
                       "Unnormalized: Mean division log10",
                       "ComBat: Mean division log10",
                       "fda: Mean division log10")
plot_values$method = method_values

## calculate mean misclassification
mean_values = data.frame(rowMeans(plot_values[,-c(1)]))
mean_values$method = method_values
colnames(mean_values)[1] = c("means")

## plotting variables
size_val = 3
alpha_val = .8

## fig2a plot
ggplot(plot_values) +
  geom_vline(xintercept=0)+
  geom_point(aes(x=Median_Cell_CD3D,y=method,color="CD3"),size=size_val,alpha=alpha_val)+
  geom_point(aes(x=Median_Cell_CD8,y=method,color="CD8"),size=size_val,alpha=alpha_val)+
  geom_point(aes(x=Median_Cell_COLLAGEN,y=method,color="COLLAGEN"),size=size_val,alpha=alpha_val)+
  geom_point(aes(x=Median_Cell_NAKATPASE,y=method,color="NAKATPASE"),size=size_val,alpha=alpha_val)+
  geom_point(aes(x=Median_Cell_PANCK,y=method,color="PANCK"),size=size_val,alpha=alpha_val)+
  geom_point(aes(x=Median_Cell_VIMENTIN,y=method,color="VIMENTIN"),size=size_val,alpha=alpha_val) +
  geom_point(aes(x=Median_Cell_SOX9,y=method,color="SOX9"),size=size_val,alpha=alpha_val) +
  geom_point(aes(x=Median_Cell_BCATENIN,y=method,color="BCATENIN"),size=size_val,alpha=alpha_val) +
  geom_point(aes(x=Median_Cell_OLFM4,y=method,color="OLFM4"),size=size_val,alpha=alpha_val) +
  geom_point(data = mean_values,aes(x=means,y=method,group=method),color="black",fill="white",shape=23,size=3)+
  scale_x_continuous("Change in mean misclassification for method, compared to raw data")+
  scale_y_discrete("")+
  scale_color_discrete("Marker") +
  theme(axis.text.y = element_text(hjust = 0))
```

### Figure 2B: Otsu misclassification (using manual labels)
```{r fig2b,fig.height=6}
## --- calculate Otsus on manual labels for CD3 & CD8
cd3s = c(); cd8s = c()
pred_methods = c(methods, "raw")

## calculate Otsu threshold
for(m in pred_methods){
  cd3s = c(cd3s,get_otsu_sk(sc[,paste0("Median_Cell_CD3D","_",m)]))
  cd8s = c(cd8s,get_otsu_sk(sc[,paste0("Median_Cell_CD8","_",m)]))
}

## generate boolean for marker positive
for(i in 1:length(cd3s)){
  m = pred_methods[i]
  sc[,paste0("CD3_Otsu_Boolean_",m)] = sc[,paste("Median_Cell_CD3D",m,sep = "_")] > cd3s[i]
  sc[,paste0("CD8_Otsu_Boolean_",m)] = sc[,paste("Median_Cell_CD8",m,sep = "_")] > cd8s[i]
}

## turn manual labels into booleans
sc$cd3_bool = ifelse(sc$CD3D=="CD3D+",1,0)
sc$cd8_bool = ifelse(sc$CD8=="CD8+",1,0)

## --- calculate accuracy of maintaining manual labels
cd3_acc = c()
cd8_acc = c()
for(m in pred_methods){
  cmat1 = caret::confusionMatrix(as.factor(sc$cd3_bool), as.factor(sc[,paste0("CD3_Otsu_Boolean_",m)]*1))
  cd3_acc = c(cd3_acc,cmat1$overall["Accuracy"]) 
  
  cmat2 = caret::confusionMatrix(as.factor(sc$cd8_bool), as.factor(sc[,paste0("CD8_Otsu_Boolean_",m)]*1))
  cd8_acc = c(cd8_acc,cmat2$overall["Accuracy"])
}

## subtract raw from each method
cd3_acc = cd3_acc - cd3_acc[13]
cd8_acc = cd8_acc - cd8_acc[13]

## create plotting dataset
acc_dat = data.frame(c(pred_methods[-13],pred_methods[-13]))
colnames(acc_dat) = c("method")
acc_dat$method = unname(unlist(adjList[acc_dat$method]))
acc_dat$values = c(cd3_acc[-13],cd8_acc[-13])
acc_dat$marker = c(rep("CD3",12),rep("CD8",12))

## fig2b plot
ggplot(acc_dat)+
  geom_vline(xintercept = 0)+
  geom_point(aes(x=values,y=method,color=marker),size=3) +
  scale_x_continuous("Change in marker-positive accuracy, compared to raw data")+
  scale_y_discrete("")+
  scale_color_manual("Marker",values=c("lightskyblue","chocolate3")) +
  theme(axis.text.y = element_text(hjust = 0),legend.position = "right")
```

### Figure 3: Proportion of variance present at slide-level in random effects model for CD3 and CD8 markers
```{r fig3}
## --- --- ---
### --- get baseline values from raw data
## --- --- ---

m = "raw"
## ---- CD3 
## modeling
cd3_form = as.formula(paste0("Median_Cell_CD3D_",m,"~epi*lessBroad + (1|SlideID)"))
cd3_mod = lmer(cd3_form,data=sc)

## setup variance vals
varcomp_cd3 = (c(unname(attr(summary(cd3_mod)$varcor$SlideID,"stddev")),
        summary(cd3_mod)$sigma))^2
varcomp_cd3 = varcomp_cd3/sum(varcomp_cd3)
svar1 = varcomp_cd3[1]

## ---- CD8
## modeling
cd8_form = as.formula(paste0("Median_Cell_CD8_",m,"~epi*lessBroad + (1|SlideID)"))
cd8_mod = lmer(cd8_form,data=sc)

## setup variance vals
varcomp_cd8 = (c(unname(attr(summary(cd8_mod)$varcor$SlideID,"stddev")),
        summary(cd8_mod)$sigma))^2
varcomp_cd8 = varcomp_cd8/sum(varcomp_cd8)
svar2 = varcomp_cd8[1]

## --- --- ---
### --- get variance components from each method
## --- --- ---
cd3_vcs = data.frame()
cd8_vcs = data.frame()

for(m in methods){
  ## ---- CD3 
  ## modeling
  cd3_form = as.formula(paste0("Median_Cell_CD3D_",m,"~epi*lessBroad + (1|SlideID)"))
  cd3_mod = lmer(cd3_form,data=sc)
  varcomp_cd3 = (c(unname(attr(summary(cd3_mod)$varcor$SlideID,"stddev")),
          summary(cd3_mod)$sigma))^2
  
  ## setup variance vals
  varcomp_cd3 = varcomp_cd3/sum(varcomp_cd3)
  varcomp_cd3 = data.frame(varcomp_cd3)
  
  ## data setup
  colnames(varcomp_cd3)[1] = c("variance")
  varcomp_cd3$level = c("Slide","Residual")
  varcomp_cd3$method = m
  cd3_vcs = rbind(cd3_vcs,varcomp_cd3)
  
  ## ---- CD8
  ## modeling
  cd8_form = as.formula(paste0("Median_Cell_CD8_",m,"~epi*lessBroad + (1|SlideID)"))
  cd8_mod = lmer(cd8_form,data=sc)
  varcomp_cd8 = (c(unname(attr(summary(cd8_mod)$varcor$SlideID,"stddev")),
          summary(cd8_mod)$sigma))^2
    
  ## setup variance vals
  varcomp_cd8 = varcomp_cd8/sum(varcomp_cd8)
  varcomp_cd8 = data.frame(varcomp_cd8)  
  
  ## data setup
  colnames(varcomp_cd8)[1] = c("variance")
  varcomp_cd8$level = c("Slide","Residual")
  varcomp_cd8$method = m
  cd8_vcs = rbind(cd8_vcs,varcomp_cd8)
}

#cd3_vcs = cd3_vcs[!(cd3_vcs$method %in% not_cols),]
#cd8_vcs = cd8_vcs[!(cd8_vcs$method %in% not_cols),]

## change values
cd3_vcs$method = unname(unlist(adjList[cd3_vcs$method]))
cd8_vcs$method = unname(unlist(adjList[cd8_vcs$method]))

## generate plots
gcd3 = ggplot(cd3_vcs) +
  geom_col(aes(y=method,x=variance,fill=level)) +
  theme(axis.text.y = element_text(hjust = 1)) +
  scale_fill_manual("Level",values=rev(viridis::magma(2)))+
  geom_vline(xintercept=svar1) +
  scale_x_continuous("Proportion of variance")+
  scale_y_discrete("")+
  ggtitle("CD3") +
  theme(axis.text.y = element_text(size = 10))

gcd8 = ggplot(cd8_vcs) +
  geom_col(aes(y=method,x=variance,fill=level)) +
  theme(axis.text.y = element_text(hjust = 1)) +
  scale_fill_manual("Level",values=rev(viridis::magma(2)))+
  geom_vline(xintercept=svar2) +
  scale_x_continuous("Proportion of variance")+
  scale_y_discrete("")+
  ggtitle("CD8") +
  theme(axis.text.y = element_text(size = 10))

## fig3 plot
ggpubr::ggarrange(gcd3,gcd8,nrow=2,ncol=1,legend = "bottom",common.legend = TRUE)
```

### Figure 4: Comparison of cell proportions for each transformation method
#### 4a: Manually labelled cells
```{r fig4a}
tts = read.csv('/media/disk2/atlas_mxif/data/TissueSubTypes Colon Map.csv')

## --- match capitalization and types, combine data
names(tts)[1] = 'TissueID'
tts$adenSSLcrc = ifelse(tts$tissueSubType %in% c('Tub', 'TV'), 'AD', 
                        ifelse(tts$tissueSubType %in% c('HP', 'HP/SSL', 'SSL', 'Tub/SSL'), 'SSL', 
                               ifelse(tts$tissueSubType %in% c('CRC'), 'CRC', NA )) )
tts$SlideID = sc$SlideID[match(tts$TissueID, sc$TissueID)]
tts = tts[,c('SlideID', 'adenSSLcrc')]
sc = merge(sc, tts, all.x=TRUE)
sc = sc[ sc$lessBroad %in% c('AD', 'SSL', 'HP'),]
sc = sc[ sc$Tumor>0,]

#sc$cd3_bool = ifelse(sc$CD3D=="CD3D+",1,0)
#sc$cd8_bool = ifelse(sc$CD8=="CD8+",1,0)

## --- group data
small_sc= sc %>% 
  group_by(Slide_Region,epi,SlideID,lessBroad,.add=TRUE) %>% 
  summarise_at(colnames(sc)[c(which(grepl("Otsu_Boolean",colnames(sc))),which(colnames(sc) %in% c("cd3_bool","cd8_bool")))],mean)

small_sc$epi= ifelse(small_sc$epi == 1,"Epithelium","Stroma")

# methods = c("log10","mean_division","mean_division_log10","cube_root","fifth_root","raw")
# methods = c(methods,paste0(methods,"_fda_registered_x1"))
# methods = c(methods,paste0(methods[1:6],"_ComBat_incl_zeroes"))
# not_cols = c("fifth_root","fifth_root_ComBat_incl_zeroes","fifth_root_fda_registered_x1",
#              "raw","raw_ComBat_incl_zeroes",
#              "raw_fda_registered_x1")
# methods = methods[!(methods %in% not_cols)]

## --- setup variables for analysis
methods = sort(methods)[c(4:6,1:3,7:9,10:12)]
methods_bool = paste0("Otsu_Boolean_",methods)
CD3_vars = paste("CD3", methods_bool, sep="_" )
CD8_vars = paste("CD8", methods_bool, sep="_" )
CD3_varcomps = data.frame()
CD8_varcomps = data.frame()

run_var_comps = function(v){
  ## --- generate model and fits
  formula_v = as.formula(paste0(v,"~epi*lessBroad+(1|SlideID)"))
  mod_v = lmer(formula_v,data=small_sc)
  CIs_v = emmeans(mod_v, as.formula(paste('~ 1 | ', "lessBroad", ' + epi')), type='response')
  fits_v = as.data.frame(summary(CIs_v))
  
  ## --- calc varcomps
  varcomp_v = (c(unname(attr(summary(mod_v)$varcor$SlideID,"stddev")),
            summary(mod_v)$sigma))^2
  varcomp_v = data.frame(varcomp_v)
  varcomp_v$level = c("Slide","Cell (Residual)")
  varcomp_v$method = stringr::str_replace(v,"CD3_Otsu_Boolean_","")
  colnames(varcomp_v)[1] = c("Variance")
  
  return(list("varcomps"=varcomp_v,
              "fits"=fits_v))
}

get_cp_plot = function(v,fits,m_v,marker="CD3",dat=small_sc){
  ggplot(dat, aes(x=as.factor(epi),y=get(v),color=as.factor(epi),fill=as.factor(epi)))+
      gghalves::geom_half_point(side = "l",range_scale = .4, alpha = .5) +
      geom_boxplot(width=0.1,alpha=0.5,outlier.shape = NA) +
      ## random effects estimated means, not included.
      # geom_jitter(data=fits[fits$lessBroad=="AD",],aes(x=as.factor(epi),y=emmean),size=2,width=0.05,shape=23,fill="lightblue4") +
      # geom_jitter(data=fits[fits$lessBroad=="HP",],aes(x=as.factor(epi),y=emmean),size=2,width=0.05,shape=23,fill="midnightblue") +
      # geom_jitter(data=fits[fits$lessBroad=="SSL",],aes(x=as.factor(epi),y=emmean), size=2,width=0.05,shape=23,fill="lightslateblue") +
      ggdist::stat_halfeye(adjust = .85, width = .5, .width = 0, justification = -.25, point_colour = NA,alpha=0.5) +
      scale_x_discrete(m_v) +
      scale_y_continuous(paste0(marker," Proportions"),limits=c(-0.05,1.05))+
      scale_fill_manual(values=c("lightslateblue","coral1"))+
      scale_color_manual(values=c("lightslateblue","coral1"))+
      theme(legend.position = "None")
      #scale_fill_manual("Tumor Type",values=viridis::magma(10)[c(4,7,10)])
}

## CD3 base
vb1 = "cd3_bool"
vbobj1 = run_var_comps(v = vb1)
b1 = get_cp_plot(vb1, vbobj1$fits,m_v="CD3 Ground Truth Label",marker="CD3")

## CD8 base
vb2 = "cd8_bool"
vbobj2 = run_var_comps(v = vb2)
b2 = get_cp_plot(vb2, vbobj2$fits,m_v="CD8 Ground Truth Label",marker="CD8")

## fig4a plot
bfig = ggpubr::ggarrange(b1,b2,common.legend = TRUE,legend = "none")
ggpubr::annotate_figure(bfig, top=text_grob("Baseline proportions in CD3 and CD8 using manual labels"))
```

#### 4b: CD3 for each method
```{r fig4b,fig.height=10}
v1 = CD3_vars[1]
vobj1 = run_var_comps(v = v1)
g1 = get_cp_plot(v1, vobj1$fits,m_v="Unnormalized: log10",marker="CD3")

v2 = CD3_vars[2]
vobj2 = run_var_comps(v = v2)
g2 = get_cp_plot(v2, vobj2$fits,m_v="ComBat: log10",marker="CD3")

v3 = CD3_vars[3]
vobj3 = run_var_comps(v = v3)
g3 = get_cp_plot(v3, vobj3$fits,m_v="fda: log10",marker="CD3")

v4 = CD3_vars[4]
vobj4 = run_var_comps(v = v4)
g4 = get_cp_plot(v4, vobj4$fits,m_v="Unnormalized: Cube root",marker="CD3")

v5 = CD3_vars[5]
vobj5 = run_var_comps(v = v5)
g5 = get_cp_plot(v5, vobj5$fits,m_v="ComBat: Cube root",marker="CD3")

v6 = CD3_vars[6]
vobj6 = run_var_comps(v = v6)
g6 = get_cp_plot(v6, vobj6$fits,m_v="fda: Cube root",marker="CD3")

v7 = CD3_vars[7]
vobj7 = run_var_comps(v = v7)
g7 = get_cp_plot(v7, vobj7$fits,m_v="Unnormalized: Mean division",marker="CD3")

v8 = CD3_vars[8]
vobj8 = run_var_comps(v = v8)
g8 = get_cp_plot(v8, vobj8$fits,m_v="ComBat: Mean division",marker="CD3")

v9 = CD3_vars[9]
vobj9 = run_var_comps(v = v9)
g9 = get_cp_plot(v9, vobj9$fits,m_v="fda: Mean division",marker="CD3")

v10 = CD3_vars[10]
vobj10 = run_var_comps(v = v10)
g10 = get_cp_plot(v10, vobj10$fits,m_v="Unnormalized: Mean division log10",marker="CD3")

v11 = CD3_vars[11]
vobj11 = run_var_comps(v = v11)
g11 = get_cp_plot(v11, vobj11$fits,m_v="ComBat: Mean division log10",marker="CD3")

v12 = CD3_vars[12]
vobj12 = run_var_comps(v = v12)
g12 = get_cp_plot(v12, vobj12$fits,m_v="fda: Mean division log10",marker="CD3")

## fig4b plot
cd3fig = ggpubr::ggarrange(plotlist=list(g1,g2,g3,
                           g4,g5,g6,
                           g7,g8,g9,
                           g10,g11,g12),common.legend = TRUE,legend = "none", ncol = 3,nrow=4)
ggpubr::annotate_figure(cd3fig, top=text_grob("Proportions of CD3 by method"))
```

#### 4c: CD8 for each method
```{r fig4c,fig.height=10}
v1 = CD8_vars[1]
vobj1 = run_var_comps(v = v1)
g1 = get_cp_plot(v1, vobj1$fits,m_v="Unnormalized: log10",marker="CD8")

v2 = CD8_vars[2]
vobj2 = run_var_comps(v = v2)
g2 = get_cp_plot(v2, vobj2$fits,m_v="ComBat: log10",marker="CD8")

v3 = CD8_vars[3]
vobj3 = run_var_comps(v = v3)
g3 = get_cp_plot(v3, vobj3$fits,m_v="fda: log10",marker="CD8")

v4 = CD8_vars[4]
vobj4 = run_var_comps(v = v4)
g4 = get_cp_plot(v4, vobj4$fits,m_v="Unnormalized: Cube root",marker="CD8")

v5 = CD8_vars[5]
vobj5 = run_var_comps(v = v5)
g5 = get_cp_plot(v5, vobj5$fits,m_v="ComBat: Cube root",marker="CD8")

v6 = CD8_vars[6]
vobj6 = run_var_comps(v = v6)
g6 = get_cp_plot(v6, vobj6$fits,m_v="fda: Cube root",marker="CD8")

v7 = CD8_vars[7]
vobj7 = run_var_comps(v = v7)
g7 = get_cp_plot(v7, vobj7$fits,m_v="Unnormalized: Mean division",marker="CD8")

v8 = CD8_vars[8]
vobj8 = run_var_comps(v = v8)
g8 = get_cp_plot(v8, vobj8$fits,m_v="ComBat: Mean division",marker="CD8")

v9 = CD8_vars[9]
vobj9 = run_var_comps(v = v9)
g9 = get_cp_plot(v9, vobj9$fits,m_v="fda: Mean division",marker="CD8")

v10 = CD8_vars[10]
vobj10 = run_var_comps(v = v10)
g10 = get_cp_plot(v10, vobj10$fits,m_v="Unnormalized: Mean division log10",marker="CD8")

v11 = CD8_vars[11]
vobj11 = run_var_comps(v = v11)
g11 = get_cp_plot(v11, vobj11$fits,m_v="ComBat: Mean division log10",marker="CD8")

v12 = CD8_vars[12]
vobj12 = run_var_comps(v = v12)
g12 = get_cp_plot(v12, vobj12$fits,m_v="fda: Mean division log10",marker="CD8")

## fig4c plot
cd8fig = ggpubr::ggarrange(plotlist=list(g1,g2,g3,
                           g4,g5,g6,
                           g7,g8,g9,
                           g10,g11,g12),common.legend = TRUE,legend = "none", ncol = 3,nrow=4)
ggpubr::annotate_figure(cd8fig, top=text_grob("Proportions of CD8 by method"))
```

### Figure 5:  UMAP embedding of data for each transformation method
Note that the UMAP algorithm is stochastic, and because of the data size, we are downsampling so these plots may not be identical to those in the paper.

```{r umap}
## --- Multivariate clustering
## create downsampled data
scs = split(sc,sc$SlideID)
ds_scs_list = lapply(X=1:length(scs),FUN = function(i){
  nr=nrow(scs[[i]])
  scs[[i]][sample(x = 1:nr,size = round(0.1 * nr)),]
})

## set vars for clustering
dsd = data.frame(data.table::rbindlist(ds_scs_list))
epi_vars = all_vars[3:6]
all_dfs = data.frame()

## run umap on all methods
for(m in methods){
  ## get umap loadings
  u1 = tumap(dsd[,paste0(epi_vars,"_",m)])
  df_add = data.frame(u1[,1],u1[,2])
  colnames(df_add) = c("U1","U2")
  
  ## add relevant columns
  df_add$method = m
  df_add$epi = dsd$epi
  df_add$tumor = dsd$Tumor
  df_add$SlideID = dsd$SlideID
  
  all_dfs = rbind(df_add,all_dfs)
}
```


#### 5ab: Raw data, colored by tissue class and slide identifier
```{r fig5ab}
m = "raw"
u1 = tumap(dsd[,paste0(epi_vars,"_",m)])
df_add = data.frame(u1[,1],u1[,2])
colnames(df_add) = c("U1","U2")

## add relevant columns
df_add$method = m
df_add$epi = dsd$epi
df_add$tumor = dsd$Tumor
df_add$SlideID = dsd$SlideID

## tissue class plot
e1 = ggplot(df_add) + 
  geom_point(aes(x=U1,y=U2,color=factor(epi)),size=0.01,alpha=0.05) +
  scale_x_continuous("Raw (by Tissue Class)") +
  scale_y_continuous("") +
  scale_color_manual("Tissue type",labels=c("Stroma","Epithelium"),values=c("tan1","skyblue4")) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(override.aes = list(size=2)))

## slide plot
s1 = ggplot(df_add) + 
  geom_point(aes(x=U1,y=U2,color=factor(SlideID)),size=0.01,alpha=0.05) +
  scale_x_continuous("Raw (by Slide)") +
  scale_y_continuous("") +
  theme_minimal() +
  theme(legend.position = "None")

## fig5ab plot
ggpubr::ggarrange(e1,s1)
```

#### 5c: Methods colored by tissue class
```{r fig5c, fig.height=10}
msplit = split(all_dfs,all_dfs$method)
msplit = msplit[c(4,5,6,1,2,3,7,8,9,10,11,12)]
#msplit = append(msplit,values="blank",after=8)

new_labs = c("Unnormalized: log10",
                       "ComBat: log10", 
                       "fda: log10",
                       "Unnormalized: Cube root",
                       "ComBat: Cube root", 
                       "fda: Cube root",
                       "Unnormalized: Mean division",
                       "ComBat: Mean division", 
                       "fda: Mean division",
                       "Unnormalized: Mean division log10",
                       "ComBat: Mean division log10",
                       "fda: Mean division log10")

## 4x3 figure of each method UMAP
plist = lapply(X=1:length(msplit),FUN=function(i){
  m = msplit[[i]]
  ggplot(m) + 
    geom_point(aes(x=U1,y=U2,color=factor(epi)),size=0.01,alpha=0.35) +
    scale_x_continuous(paste0(new_labs[i],""))+
    scale_y_continuous("") +
    scale_color_manual("Tissue type",labels=c("Stroma","Epithelium"),values=c("tan1","skyblue4")) +
    theme_minimal() +
    guides(colour = guide_legend(override.aes = list(size=2)))
})

## fig5c plot
ggpubr::ggarrange(plotlist=plist,legend = "bottom",common.legend = TRUE,nrow=4,ncol=3)
```

#### 5d: Methods colored by slide identifier
```{r fig5d, fig.height=10}
## 4x3 figure of each method UMAP
plist = lapply(X=1:length(msplit),FUN=function(i){
  m = msplit[[i]]
    ggplot(m)  + 
      geom_point(aes(x=U1,y=U2,color=factor(SlideID)),size=0.01,alpha=0.05) +
      scale_x_continuous(paste0(new_labs[i]," (by Slide)"))+
      scale_y_continuous("") + 
      #scale_color_manual(values=vals) +
      theme_minimal() +
      theme(legend.position = "None")
  # }
})

## fig5d plot
ggpubr::ggarrange(plotlist=plist,nrow=4,ncol=3)
```

## `R` Version info
```{r}
sessionInfo()
```

